---
title: "Aggregating Provider Data for Analysis"
author:
  - Tyler Muffly, MD
output:
  rmarkdown::html_vignette:
    df_print: kable
description: >
  Build an analysis-ready provider table by combining roster, enrichment,
  geography, and call-outcome data.
vignette: >
  %\VignetteIndexEntry{Aggregating Provider Data for Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
eval_vignettes <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  eval = eval_vignettes
)

if (eval_vignettes) {
  library(mysterycall)
}
```

# Overview

Most mystery caller projects do not end with a provider roster. The analytical
work starts when you combine several layers of information into one table:

- provider identity and specialty
- validated NPI values
- clinician enrichment from external sources
- geography and travel-time metrics
- call outcomes and payer-specific access measures

This vignette outlines a practical structure for that aggregation step.

## Preserve a stable key

Use `npi` as the primary key whenever it is available. Names and addresses can
change; NPI is usually the most stable identifier across roster creation,
enrichment, and call outcome files.

```{r eval=FALSE}
roster <- search_by_taxonomy(
  "Gynecologic Oncology",
  states = c("CO", "WY"),
  write_snapshot = FALSE,
  notify = FALSE
) |>
  validate_and_remove_invalid_npi()
```

## Add clinician enrichment

Once you have a validated roster, enrich it with clinician-level attributes
such as demographics or training metadata. In the current API, the exported
function for this step is `retrieve_clinician_data()`.

```{r eval=FALSE}
clinician_data <- retrieve_clinician_data(roster)

analysis_base <- roster |>
  dplyr::left_join(
    clinician_data,
    by = "npi",
    suffix = c("_roster", "_clinician")
  )
```

## Keep site-level geography separate from clinician-level attributes

Geocoding and isochrone generation happen at the practice-site level, not
strictly at the person level. If multiple clinicians share one address, it is
more efficient to geocode the unique site once and join the coordinates back.

```{r eval=FALSE}
site_table <- analysis_base |>
  dplyr::distinct(npi, address, .keep_all = TRUE)
```

After geocoding and travel-time modelling, join the geography back onto the
analysis table using the columns you preserved upstream.

```{r eval=FALSE}
analysis_with_geo <- analysis_base |>
  dplyr::left_join(
    site_geography,
    by = c("npi", "address")
  )
```

## Join call outcomes late

Operational call files often contain repeated contacts, reschedules, or payer
scenarios. It is usually easier to clean those files first and join them late
in the pipeline.

```{r eval=FALSE}
phase2_clean <- clean_phase_2_data("phase2_results.xlsx")

analysis_ready <- analysis_with_geo |>
  dplyr::left_join(
    phase2_clean,
    by = c("npi", "name")
  )
```

## Example aggregation patterns

Once the core joins are complete, the same table can support several summary
layers.

### Count providers by subspecialty

```{r}
tyler::physicians |>
  dplyr::count(subspecialty, sort = TRUE)
```

### Summarize access outcomes

```{r eval=FALSE}
analysis_ready |>
  dplyr::count(subspecialty, medicaid_accepted)
```

### Build publication tables

```{r eval=FALSE}
table_generate_overall(
  analysis_ready,
  vars = c("subspecialty", "medicaid_accepted", "appointment_offered")
)
```

## Recommended table structure

In practice, an analysis-ready table works best when it has one row per unit of
analysis. Decide that unit explicitly:

- one row per clinician
- one row per clinician-site pair
- one row per call attempt

Do not mix these units casually in the same final table or downstream
percentages will be hard to interpret.

## Practical guidance

- Keep raw roster creation outputs separate from analysis-ready outputs.
- Join on `npi` whenever possible.
- Distinguish clinician-level and site-level variables.
- Clean repeated call attempts before calculating proportions.
- Save intermediate tables so you can audit merges and row counts.

# Next step

Once the unified table exists, the remaining tyler helpers for manuscript
outputs are mostly summarization and visualization: `table_generate_overall()`,
`table_calculate_percentages()`, and the mapping functions for geographic
presentation.
