---
title: "Aggregating Provider Data for Analysis"
author:
  - Tyler Muffly, MD
output:
  rmarkdown::html_vignette:
    df_print: kable
description: >
  A skeleton vignette describing how to aggregate provider data from multiple sources.
vignette: >
  %\VignetteIndexEntry{Aggregating Provider Data for Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

# Objectives

- Demonstrate a workflow for combining provider datasets.
- Show how to clean and standardize fields prior to merging.
- Provide a starting point for further analysis.

# Required Packages

```{r setup, include=FALSE}
eval_vignettes <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  eval = eval_vignettes
)

if (eval_vignettes) {
  library(tyler)
}
```

# Inputs

Describe the input files such as NPI exports, Census data, and geocoded locations.

```{r}
# Example placeholder for reading data
# npi_data <- readr::read_csv("npi.csv")
# census_data <- readr::read_csv("census.csv")
```

# Processing Steps

Outline the steps needed to clean names, join datasets, and summarize counts.

```{r}
# Example data manipulation steps
# combined <- dplyr::left_join(npi_data, census_data, by = "geoid")
```

# Output Examples

Show an example of the aggregated output such as counts by county.

```{r}
# aggregated <- combined %>% dplyr::count(county)
# aggregated
```

# Conclusions

Summarize the results and next steps for analysis.

# Features and bugs
If you have ideas for other features that would make data aggregation easier, or find a bug, the best approach is to either report it or add it!
