---
title: "Create Drive-Time Isochrones"
author:
  - Tyler Muffly, MD
output:
  rmarkdown::html_vignette:
    df_print: kable
description: >
  Generate HERE drive-time isochrones from geocoded practice coordinates.
vignette: >
  %\VignetteIndexEntry{Create Drive-Time Isochrones}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
eval_vignettes <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  eval = eval_vignettes
)

if (eval_vignettes) {
  library(tyler)
}
```

# Overview

`create_isochrones_for_dataframe()` builds drive-time polygons from geocoded
practice locations using the HERE API. The function is designed for batch
processing: it reads a file of point locations, validates the coordinates,
requests isochrones for each row, and periodically writes checkpoint files to
disk while it runs.

## Before you start

You will need:

- A HERE API key.
- An input file with `lat` and `long` columns in decimal degrees.
- Coordinates in WGS84 longitude/latitude space.

Store the key in `HERE_API_KEY` or pass it directly.

```{r eval=FALSE}
Sys.setenv(HERE_API_KEY = "your-here-key")
```

## Input requirements

The current exported function expects a file path via `input_file`. The file is
read with `easyr::read.any()`, so CSV, Excel, and several other tabular formats
are supported. The important requirement is that the data include:

- `lat`
- `long`

Rows outside valid WGS84 bounds are rejected before any API request is made.

## Minimal workflow

The typical pattern is:

1. Geocode practice addresses.
2. Rename the coordinate columns to `lat` and `long`.
3. Write the result to a file.
4. Pass that file into `create_isochrones_for_dataframe()`.

```{r eval=FALSE}
iso_input <- geocoded |>
  dplyr::rename(
    lat = latitude,
    long = longitude
  )

input_csv <- tempfile(fileext = ".csv")
readr::write_csv(iso_input, input_csv)

isochrones <- create_isochrones_for_dataframe(
  input_file = input_csv,
  breaks = c(30 * 60, 60 * 60, 120 * 60, 180 * 60),
  api_key = Sys.getenv("HERE_API_KEY"),
  output_dir = tempfile(),
  save_interval = 240
)
```

## Understanding the arguments

- `breaks`: travel-time cut points in seconds.
- `api_key`: HERE API key. Defaults to `HERE_API_KEY`.
- `output_dir`: directory for intermediate checkpoint files.
- `save_interval`: how often checkpoint files should be refreshed during long
  runs.

For example, `c(1800, 3600)` means 30-minute and 60-minute drive-time
isochrones.

## What gets returned

The function returns a combined `sf` object containing every generated
isochrone. It also keeps the non-geometry attributes from the source row so
that the polygons remain traceable back to the originating practice site.

Useful added columns include:

- `point_index`
- `travel_time_minutes`
- `name`
- `column_label`

```{r eval=FALSE}
names(isochrones)

isochrones |>
  dplyr::count(travel_time_minutes)
```

## Checkpoint files

Long runs can take time, so the function writes intermediate results to
`output_dir` as both:

- `.rds`
- `.gpkg`

That design is intentional. If a long run is interrupted, you still have
recently saved spatial output on disk rather than losing the entire batch.

## Practical guidance

- Use street-level geocoding first; do not build isochrones from ZIP-code
  centroids when practice-level access matters.
- Start with a small subset of rows to confirm your pipeline and API key.
- Keep `output_dir` for auditing and recovery on long production runs.
- Treat `breaks` as a design decision tied to your study question. Common
  cutoffs are 30, 60, 120, and 180 minutes.

# Next step

After generating isochrones, a common next stage is
`calculate_intersection_overlap_and_save()` to overlay Census block groups and
summarize the reachable population inside each travel-time band.
