---
title: "Geocoding Practice Addresses"
author:
  - Tyler Muffly, MD
output:
  rmarkdown::html_vignette:
    df_print: kable
description: >
  Geocode provider practice addresses with the current file-based tyler API.
vignette: >
  %\VignetteIndexEntry{Geocoding Practice Addresses}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
eval_vignettes <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  eval = eval_vignettes
)

if (eval_vignettes) {
  library(tyler)
}
```

# Overview

`geocode_unique_addresses()` geocodes one row per unique address and then joins
the coordinates back onto the original dataset. This is useful after you have
assembled a provider roster and standardized an `address` column for each
practice site.

The current interface is file-based:

- Input must be a CSV, RDS, or XLSX file.
- The file must contain an `address` column.
- Output includes new `latitude` and `longitude` columns.

## Before you start

You will need:

- A Google Maps API key with geocoding access.
- A dataset with a fully formatted street address column named `address`.
- Enough detail in the address to geocode the location directly. Street
  addresses are better than ZIP-code centroids for access studies.

Store the API key in `GOOGLE_API_KEY` or pass it directly to the function.

```{r eval=FALSE}
Sys.setenv(GOOGLE_API_KEY = "your-google-maps-key")
```

## Minimal workflow

If your roster is already in memory, write it to a temporary CSV first. That
matches the current exported API and keeps the example aligned with the
function signature.

```{r eval=FALSE}
addresses <- tibble::tibble(
  address = c(
    "1960 Ogden St, Denver, CO 80218",
    "12605 E 16th Ave, Aurora, CO 80045"
  ),
  site_name = c("Downtown clinic", "Academic center")
)

input_csv <- tempfile(fileext = ".csv")
output_csv <- tempfile(fileext = ".csv")
failed_csv <- tempfile(fileext = ".csv")

readr::write_csv(addresses, input_csv)

geocoded <- geocode_unique_addresses(
  file_path = input_csv,
  google_maps_api_key = Sys.getenv("GOOGLE_API_KEY"),
  output_file_path = output_csv,
  failed_output_path = failed_csv,
  notify = FALSE
)
```

## What the function does

`geocode_unique_addresses()`:

1. Reads the input file.
2. Deduplicates the `address` column before calling the Google API.
3. Retries failed requests up to three times.
4. Adds `latitude` and `longitude` to the unique-address table.
5. Joins those coordinates back to the full input dataset.

That behavior matters for mystery caller studies because rosters often contain
many repeated addresses for different clinicians at the same site.

## Working with the results

The returned object is a data frame with the original columns plus coordinates.
If you requested an `output_file_path`, the same data is also written to disk.

```{r eval=FALSE}
dplyr::glimpse(geocoded)

geocoded |>
  dplyr::count(is.na(latitude), is.na(longitude))
```

If you plan to create drive-time isochrones next, convert the resulting
coordinate columns into the names expected by
`create_isochrones_for_dataframe()`.

```{r eval=FALSE}
iso_input <- geocoded |>
  dplyr::rename(
    lat = latitude,
    long = longitude
  )
```

## Capturing failures

When `failed_output_path` is supplied, rows that still fail after all retries
are written to a separate file. That file is useful for manual cleanup of bad
addresses, suite numbers, or formatting problems.

```{r eval=FALSE}
failed_rows <- readr::read_csv(failed_csv, show_col_types = FALSE)
failed_rows
```

## Practical guidance

- Geocode the most specific address available, not a city or ZIP code alone.
- Standardize addresses before geocoding if they came from multiple upstream
  sources.
- Save failed rows and review them before dropping them from an analysis.
- Rename `latitude` and `longitude` to `lat` and `long` if the next step is
  `create_isochrones_for_dataframe()`.

# Next step

After geocoding, a common next stage is drive-time modelling with
`create_isochrones_for_dataframe()`, followed by Census overlap calculations
and manuscript tables.
