---
title: "Search the NPI Registry by Taxonomy"
author:
  - Tyler Muffly, MD
output:
  rmarkdown::html_vignette:
    df_print: kable
pkgdown:
  as_is: true
description: >
  Find providers in the NPI registry using taxonomy descriptions and the
  built-in taxonomy crosswalk.
vignette: >
  %\VignetteIndexEntry{Search the NPI Registry by Taxonomy}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
eval_vignettes <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = eval_vignettes
)

if (eval_vignettes) {
  library(tyler)
}
```

# Overview

`search_by_taxonomy()` is the quickest way to build a specialty-specific roster
from the National Provider Identifier registry when you know the taxonomy
description you want to target.

The package also includes a built-in `taxonomy` dataset that helps you inspect
taxonomy labels before running the search.

## Explore the built-in taxonomy crosswalk

If you are not sure which specialty strings to search, start by filtering
`tyler::taxonomy`.

```{r}
tyler::taxonomy |>
  dplyr::filter(stringr::str_detect(
    Classification,
    stringr::fixed("GYN", ignore_case = TRUE)
  )) |>
  dplyr::select(Code, Classification, Specialization) |>
  dplyr::slice_head(n = 10)
```

This step is helpful because the NPI API works with taxonomy descriptions, not
just taxonomy codes.

## Run a focused taxonomy search

The simplest call is a national search for one description.

```{r eval=FALSE}
gyn_onc <- search_by_taxonomy(
  taxonomy_to_search = "Gynecologic Oncology",
  write_snapshot = FALSE,
  notify = FALSE
)
```

## Bypass the 1,200-record cap

The NPI API limits each query to 1,200 records. For larger specialties, pass a
vector of state abbreviations. `search_by_taxonomy()` will loop over states and
combine the results.

```{r eval=FALSE}
mountain_west <- c("CO", "UT", "WY", "NM", "AZ", "ID", "MT", "NV")

fpmrs <- search_by_taxonomy(
  taxonomy_to_search = "Female Pelvic Medicine and Reconstructive Surgery",
  states = mountain_west,
  write_snapshot = FALSE,
  notify = FALSE
)
```

## What comes back

The returned data frame includes one row per matched provider-taxonomy result
after filtering to individual U.S. clinicians and removing noisy columns from
the raw API response.

Useful columns typically include:

- `npi`
- `first_name`
- `last_name`
- `addresses_city`
- `addresses_state`
- `taxonomies_desc`
- `search_term`

```{r eval=FALSE}
dplyr::glimpse(fpmrs)
```

## Common post-processing steps

After a taxonomy search, the usual next actions are:

1. Validate the NPIs.
2. Enrich the roster with clinician-level metadata.
3. Geocode the practice addresses.

```{r eval=FALSE}
validated <- validate_and_remove_invalid_npi(fpmrs)

clinician_data <- retrieve_clinician_data(validated)
```

## Practical guidance

- Search one taxonomy description at a time when you want clearer QA.
- Use state loops for large specialties to avoid truncation at 1,200 records.
- Keep `write_snapshot = TRUE` for production work if you want an audit trail of
  the retrieved roster.
- Validate NPIs before joining to other external datasets.

# Next step

If your project starts from names rather than taxonomy descriptions, see the
`Search and Process NPI Numbers` vignette for the companion workflow based on
`search_and_process_npi()`.
