---
title: "Search and Process NPI Numbers"
author:
  - Tyler Muffly, MD
output:
  rmarkdown::html_vignette:
    df_print: kable
description: >
  Search the NPI registry from clinician first and last names using the current
  tyler interface.
vignette: >
  %\VignetteIndexEntry{Search and Process NPI Numbers}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
eval_vignettes <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
library(magrittr)
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#>",
  collapse = TRUE,
  eval = eval_vignettes
)

if (eval_vignettes) {
  library(tyler)
}
```

# Overview

`search_and_process_npi()` is the name-based entry point for roster creation.
Use it when you already have clinician first and last names from a patient
directory, credentialing list, or manually curated study roster and need to
recover likely NPI matches.

The current exported interface expects a data frame passed through the `data`
argument. That data frame must contain:

- `first`
- `last`

## Minimal input

```{r}
sample_names <- tibble::tibble(
  first = c("Jane", "Maria", "Alex"),
  last = c("Smith", "Garcia", "Johnson")
)

sample_names
```

## Basic search

The simplest call looks like this:

```{r eval=FALSE}
results <- search_and_process_npi(
  data = sample_names,
  notify = FALSE
)
```

By default the function searches for U.S. individual NPIs and filters to common
physician credentials such as `MD` and `DO`.

## Customizing the search

For most projects you will want to set a few operational parameters explicitly,
especially if the roster is large.

```{r eval=FALSE}
results <- search_and_process_npi(
  data = sample_names,
  enumeration_type = "ind",
  limit = 10,
  country_code = "US",
  filter_credentials = c("MD", "DO"),
  notify = FALSE
)
```

## Tracking long runs

Large name-based searches can take time. The function supports progress logs,
chunked saves, and resumable accumulation files.

```{r eval=FALSE}
results <- search_and_process_npi(
  data = sample_names,
  limit = 10,
  dest_dir = "npi_chunks",
  accumulate_path = "npi_results.csv",
  resume = TRUE,
  progress_log = "npi_progress.csv",
  progress_log_format = "csv",
  heartbeat_seconds = 30,
  notify = FALSE
)
```

This is useful when the roster is large enough that you want durable progress
tracking or expect to resume after an interruption.

## What comes back

The returned object is a data frame of candidate NPI matches. Typical columns
include provider identity, credentials, location fields, taxonomy descriptions,
and a `search_term` column showing the input name pair that produced the match.

```{r eval=FALSE}
dplyr::glimpse(results)

results |>
  dplyr::count(search_term, sort = TRUE)
```

## Cleaning the result set

Name-based matches are often one-to-many. A common first cleanup step is to
deduplicate on `npi` and then inspect specialty or geography fields.

```{r eval=FALSE}
results_unique <- results |>
  dplyr::distinct(npi, .keep_all = TRUE)

results_unique |>
  dplyr::count(taxonomies_desc, sort = TRUE)
```

If the next step is clinician enrichment, pass the cleaned result into
`retrieve_clinician_data()`.

```{r eval=FALSE}
clinician_data <- retrieve_clinician_data(results_unique)
```

## Practical guidance

- Start with a small roster to confirm that your filters are behaving as
  expected.
- Expect ambiguous matches for common names.
- Save progress when running large overnight jobs.
- Deduplicate and validate before downstream joins.

# Next step

If your project begins from a specialty rather than a name list, see the
taxonomy-search vignette for `search_by_taxonomy()`.
